Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-04-04 paper.tex
theory+results updated

%NEW
\newcommand{\rmn}{\mathrm}
\newcommand{\bx}{{\bf x}}
\newcommand{\fg}{{F_\gamma}}
\newcommand{\sg}{{S_\gamma}}
\newcommand{\psf}{\theta_\rmn{res}}
\newcommand{\ph}{\rmn{ph}}
\newcommand{\eph}{E_\ph}
\newcommand{\nph}{N_\ph}
\newcommand{\vir}{\rmn{vir}}
\newcommand{\gal}{\rmn{gal}}
\newcommand{\sd}{\rmn{SD}}
\newcommand{\clu}{\rmn{cl}}
\newcommand{\sfe}{\rmn{sfe}}
\newcommand{\sub}{\rmn{sub}}
\newcommand{\msun}{M_\odot}
\newcommand{\ee}{E_\rmn{e}}
\newcommand{\stars}{\rmn{stars}}
\newcommand{\dust}{\rmn{dust}}
\newcommand{\lx}{L_\rmn{X}}
\newcommand{\s}{\rmn{s}}
\newcommand{\Kp}{\rmn{K}`}
\newcommand{\Ip}{\rmn{I}`}
\newcommand{\Js}{\rmn{J}^*}
\newcommand{\Jp}{\rmn{J}`}

%OLD
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\gev}{\rm{GeV}}
\newcommand{\tev}{\rm{TeV}}
\newcommand{\ev}{\rm{eV}}
\newcommand{\dd}{\rmn{d}}
\newcommand{\mx}{\ensuremath{m_{\chi}}}
\newcommand{\ga}{\gamma}
\newcommand{\ngamma}{\ensuremath{N_{\gamma}}}
\newcommand{\ngammaj}{\ensuremath{N_{\gamma,j}}}
\newcommand{\sigmaannv}{\ensuremath{\langle\sigma v\rangle}}
\newcommand{\sigv}{\ensuremath{\sigma_v}}
\newcommand{\ic}{\rmn{IC}}
\newcommand{\egamma}{\ensuremath{E_{\gamma}}}
\newcommand{\dic}{\rmn{dIC}}
\newcommand{\fsr}{\rmn{FSR}}
\newcommand{\CR}{\rmn{CR}}
\newcommand{\rhos}{\ensuremath{\rho_s}}
\newcommand{\rs}{\ensuremath{r_s}}
\newcommand{\rvir}{r_{200}}
\newcommand{\mvir}{M_{200}}
\newcommand{\rhoc}{\ensuremath{\rho_c}}
\newcommand{\B}{\rmn{B}}
\newcommand{\e}{\rmn{e}}
\newcommand{\eg}{E_\gamma}
\newcommand{\ir}{\rmn{ir}}

@@

Among the astrophysical systems which will be very interesting to
detect, and study, with coming gamma-ray detectors (Fermi, HESS,
MAGIC, VERITAS, and eventually large detectors like CTA) belong galaxy
clusters. The most promising directions in which to search for a
gamma-ray annihilation signal (from the annihilation process itself,
and also the accompanying bremsstrahlung and inverse Compton
components coming from charged particles produce in the annihilations)
are basically three:

@@

3. Galaxy clusters. This possibility has been less studied, however we
noted in a previous Letter \cite{2009PhRvL.103r1302P} that the are
certain advantages that work in favour of this possible target for
gamma-ray detection of dark matter annihilation.  Galaxy clusters
constitute the most massive objects in our Universe that are forming
today. This causes their DM subhalo mass function to be less affected
by tidal stripping compared to galaxy sized halos that formed long
ago.  The annihilation luminosities of the DM halo component for
e.g. the Virgo cluster and the Draco dwarf scales in a way (see
\cite{2009PhRvL.103r1302P}) that the ratio of gamma-ray luminosities
from the smooth components is around 4, in favour of Virgo. In
addition, there may be a further enhancement due to substructure,
which to a large extent should be unaffected by tidal disruption, at
least in the outer regions.  According to a recent estimate
\cite{2011MNRAS.410.2309G}, more massive haloes tend to have a larger
mass fraction in subhalos.  For example, cluster size haloes typically
have 7.5 per cent of the mass within $r_{200}$ in substructures of
fractional mass larger than $10^{-5}$, which is 25 per cent higher
than galactic haloes \footnote{All halo masses and length scales in
  this paper are scaled to the currently favored value of Hubble's
  constant, $H_0 = 70\, \rm{km~s}^{-1}\,\rm{Mpc}^{-1}$. We define the
  virial mass $\mvir$ and virial radius $\rvir$ as the mass and radius
  of a sphere enclosing a mean density that is 200 times the critical
  density of the Universe $\rho_{\rm{cr}}$.}.

@@

@@

that the radius of the source $r=\sqrt{l^2+D^2-2Dl\cos\Psi}$, where

@@

We denote the total source function by $q_{\rm sum}(\eg,r)$, where we
include contributions from five main processes; leptophilic DM
annihilating to $\mu^+/\mu^-$ decaying to $\rm{e}^+/\rm{e}^-$ pairs
that up-scatter background photons (IC), leptophilic DM emitting final
state radiation (FSR), DM benchmark models where neutralino
annihilations generate electrons that up-scatter background photons
(BM-IC) and emit a continuum as well as final state radiation
(BM-Cont), and CR induced $\pi^0$:s decaying into gamma-rays. The
source function is given by

@@

q_\rmn{sum} (\eg,r) = q_{\pi^0-\gamma}(\eg,r)+

@@

where the subscript $i$ runs over the gamma-ray producing DM
channels. The differential CR to gamma-ray source function is denoted
by $q_{\pi^0-\gamma}(\eg,r)$, where we refer the reader to
Sect.~\ref{sect:CRs} and \cite{2010MNRAS.409..449P} for further
details. The total differential boost factor for DM is the product of
enhancement factors from SFE $\B_\sfe(\sigv)$ (see
Sect.~\ref{sect:LP}) and from substructure enhancement over the smooth
halo contribution $\B_{\sub,i}(r) = 1+q_\sub(r)/q_{\rmn{sm},i}(r)$
(see Sect.~\ref{sect:subst}) and is denoted by
$\B_{\rmn{tot},i}(r,\sigv) = \B_\sfe(\sigv)\,\B_{\sub,i}(r)$. The DM
source function from the smooth halo for each process is written on
the form:

@@

\frac{\dd \ngammaj}{\dd E_\gamma} \Gamma_j(r)\,.

@@

Here the annihilation rate density is given by 

@@

\Gamma_j(r) = \frac{1}{2} \left(\frac{\rho(r)}{\mx}\right)^2 
\, \sigmaannv_j\,

@@

where the subscript $j$ runs over all kinematically allowed gamma-ray

@@

clusters. It can be considered as a special case of the 5-parameter profile:

@@

and the characteristic density by 

@@

@@

Navarro-Frenk-White (NFW) profile, but provide a better fit to
simulated DM halos \cite{2010MNRAS.402...21N}. We denote the density
where the profile has a slope of $-2$ by $\rho_{-2}$, and the radius
by $r_{-2}$. Assuming that all the flux from an NFW profile originate
from within the scale radius $\rs=\rvir/c$, and all the flux from an
Einasto profile originate within $r_{-2}=\rvir/c$, it follows that
$\rho_{-2} = \rhos/4$. The virial radius $\rvir$ of a halo is defined
to be the radius at which the mean density within is a factor
$\Delta=200$ times the critical density $\rhoc$ of the universe.

\subsubsection{Substructures}
\label{sect:subst}
High-resolution DM only simulations of Milky Way (MW) type halos find
substantial amount of substructures in the periphery of DM halos,
while the substructures in the center are erased due to dynamical
disruption. Since the rate of which DM is annihilating depends on the
density squared, the resulting flux is boosted compared to the smooth
density distribution. Recent simulations suggest a flux enhancement
from substructures of the order of ten up to a few 100
\cite{2008MNRAS.391.1685S, 2008JPhCS.125a2008K} for a MW size
halo. Using the Aquarius simulations \cite{2008Natur.456...73S}, we
fit \footnote{Our approach of fitting the scaling behavior of
  $\rmn{L_{sub}}(<r)$ directly from numerical simulations
  self-consistently accounts for the radial dependence of the
  substructure concentration \cite{2008MNRAS.391.1685S}.} to the
luminosity from smooth substructures (i.e. substructures within
substructures are not included) inside radius $r$. The best fit is
given by
\begin{equation}
  {\rm L_{sub}}(<r)= 0.76\,C_1\,C_2(\mvir)\,{\rm L_{200sm}}\,x^{0.95
    \displaystyle{x}^{-0.27}}\,,
  \label{eq:LSub}
\end{equation}
where ${\rm L_{200sm}}$ is the luminosity from the smooth halo without
substructures within $\rvir$ and $x=
\left(\frac{r}{\rvir}\right)$. The first normalization constant is
derived from the simulations:
\begin{equation}
  C_1=\left(\frac{M_{\rm res,sim}}{M_{\rm lim}}\right)^{0.226}\,,
   \label{eq:C1}
\end{equation}
where $M_{\rm res,sim}=10^5\msun$ is the minimum subhalo mass in the
MW simulation, and $M_{\rm lim}$ the free streaming mass. Here we
assume that the power-law scaling relation in Eq.~(\ref{eq:C1}) is
valid down to the minimum mass of DM halos, which is conventionally in
the cold dark matter universe
\cite{2001PhRvD..64h3507H,2005JCAP...08..003G} taken to be
$10^{-6}\msun$. Note that potentially the power-law could flatten
towards smaller mass scales although current simulations show no hits
of such a behavior and in addition we are approaching the asymptotic
behavior in the power spectrum on these scales(CHECK). For DM halos
more (less) massive than the MW we expect a larger (smaller) boost
from substructures, simply because of the larger (smaller) mass range
down to the minimum mass $M_\rmn{lim}$. We capture this halo mass
dependence with the second normalization in Eq.~(\ref{eq:LSub}):
\begin{equation}
C_2(\mvir)=\left(\frac{\mvir}{M_{\rm 200sim}}\right)^{0.226}\,,
 \label{eq:C2}
\end{equation}
where $M_{\rm 200sim}=1.9\times10^{12}\msun$ is the mass of the MW
halo in the simulation \cite{2008Natur.456...73S}.

Recent detailed galaxy cluster observations find a core like density
profile with an inner slope of $\beta=0.6$, where $\beta\gtrsim 1$ can
be ruled out to a $>95$\% confidence \cite{2011ApJ...728L..39N}. This
finding is hard to reconcile with simulations that typically prefere a
steeper central profile. The different density profiles have a large
impact on the luminosity from annihilating DM, although the details of
the density profile can be neglected compared to the dominating boost
from substructures. In Fig.~\ref{fig:radial_lum} we compare the radial
dependence of the accumulative luminosity from different smooth
cluster density profiles to the substructure boosted luminosity at
different mass scales. We recalculate the density contrast for the
core like density profile $\rhos$ and rescale the concentration
parameter in Eq.~(\ref{eq:cfit}) with $300/160$
\cite{2011ApJ...728L..39N} to account for the more centralized scale
radius in a core like cluster. The emission coming from the core
compared to a cuspy like density profile is about 30\% larger within
$\rvir$. This difference is build up within 10\% of $\rvir$
(i.e. close to $r_\s$). The emission from an Einasto density
profile is about 50\% larger than the cuspy NFW profile in the
periphery of the cluster, where the difference is mainly build up at a
few percent of $\rvir$. The substructure boosted luminosities is
neglible in the center of halos, but integrated out to $\rvir$ the
boost is about; 10 for dwarf galaxies, $10^2$ for galaxies, and $10^3$
for galaxy clusters.
\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/dens.prof.eps}
\caption{\it Radial dependence of various luminosity
  contributions. The solid lines show the accumulative smooth
  luminosity from a cluster with the mass $\mvir=10^{15}\,\msun$ for
  three different density profiles; an Einasto profile with
  $\alpha=0.17$ (light blue), a cuspy NFW profile with $\beta=1.0$
  (dark blue), and a core NFW profile with $\beta=0.6$ (purple). The
  dashed lines show the accumulative luminosity from substructures for
  three different mass scales; an $\mvir=10^{15}\,\msun$ galaxy
  cluster (red), an $\mvir=10^{12}\,\msun$ galaxy (orange), and an
  $\mvir=10^{8}\,\msun$ dwarf galaxy (green). All luminosities have
  been normalized with the luminosity within $\rvir$ from a cuspy NFW
  profile. We have assumed the standard value for the limiting
  substructure mass of $10^{-6}\,\msun$. Note the large expected
  boost from substructures in clusters ($\sim1000$), and the relative
  small boost in dwarf galaxies ($\sim20$).}
 \label{fig:radial_lum}
\end{figure}

Even though the bulk of substructures have been erased in the central
regions of DM halos, a cluster in projection has a significant
enhancement due to substructures at a radius of just a few percent of
$\rvir$. In Fig.~\ref{fig:radial_emis} we show the luminosity
dominating radius for substructures and smooth density profiles, both
for three different mass scales. It clearly shows how that the main
contribution from the substructures is coming from outer parts of the
DM halos, while the main contribution from the smooth density profiles
peaks around the scale radius. NOT QUITE, ALSO FIGURE IS SHOWING DIFF
VOLUME, FAC 3 DIFFERENCE, WHAT DO WE REALLY LEARN FROM THE FIGURE
\begin{figure}
  \includegraphics[width=0.99\columnwidth]{figures/emissiv.sub.eps}
  \caption{\it Radial dependence of differential luminosity at
    different mass scales. The solid lines show the emissivity from a
    cuspy $\beta=1.0$ NFW density profile for three different mass
    scales; an $\mvir=10^{15}\,\msun$ galaxy cluster (red), an
    $\mvir=10^{12}\,\msun$ galaxy (orange), and an
    $\mvir=10^{8}\,\msun$ dwarf galaxy (green). The dashed lines show
    the emissivity from substructures for the same three mass
    scales. All emissivities have been normalized with the luminosity
    within $\rvir$ from a cuspy NFW profile. We have assumed the
    standard value for the limiting substructure mass of
    $10^{-6}\,\msun$.}
  \label{fig:radial_emis}
\end{figure}

\subsubsection{Cosmic-ray induced gamma-ray emission}
\label{sect:CRs}
The formation process of a galaxy cluster is a very energetic
processes that induces both turbulence as well as frequently occurring
merging and accretion shocks. These processes are thought to
accelerate large quantities of relativistic non-thermal electrons and
protons to high energies. On smaller scales, such as galaxies, there
is convincing evidence of such non-thermal populations. Especially, in
the MW, the cosmic rays are observed directly as well as indirectly
through radio, X-ray, and gamma-ray emission. On larger scales of the
order of few 100~kpc up to Mpc, there are a vast number of
observations of radio emission coming from radio mini halos in the
center of cooling flow clusters, radio relics in the periphery of
clusters \cite{2004rcfg.proc..335K} as well smooth giant radio halos
that have been observed in more than 50 clusters
\cite{2003ASPC..301..143F,2008SSRv..134...93F}. Most of this observed
radio emission is expected to emerge from synchrotron emitting
cosmic-ray electrons whose precise origin in especially relics and
giant radio halos is still not settled. However, due to the smoothness
of the extended emission and the observed power-law slope it seems
quite natural that they are of hadronic origin. Especially since the
long cooling time of cosmic ray protons (CRs) allows for both an
extended and large population of CRs to build up over the cluster
history \cite{1997ApJ...487..529B}.

When the CRs interact with ambient gas protons, it results in the
production of both charged and neutral pions that decay into
electrons, neutrinos, and gamma-rays. The production of these
secondaries depend both on the gas and CR densities in the cluster,
where the CR density roughly trace the gas outside the core regime and
is slightly enhance in the center. This density scaling implies that
clusters are great targets for Cherenkov telescopes with a high
sensitivity for the central parts of close by clusters. Detecting the
cluster gamma-ray emission is crucial in this respect as it
potentially provides the unique and unambiguous evidence of CR
populations in clusters through observing the $\pi^0$ bump at about
100~MeV in the spectra.

We adopt the spectrally and spatially universal gamma-ray model
developed by Pinzke \& Pfrommer \cite{2010MNRAS.409..449P} to estimate
the emission from decaying $\pi^0$:s that dominates over the inverse
Compton emission from primary and secondary electrons above 100~MeV in
clusters. The gamma-ray formalism was derived from high resolution
simulations of clusters of galaxies that included radiative
hydrodynamics, star formation, supernova feedback, and followed CR
physics using a novel formulation that trace the most important
injection and loss processes self-consistently while accounting for
the CR pressure in the equation of motion
\cite{2008A&A...481...33J,2007A&A...473...41E,2006MNRAS.367..113P}.
We note that the overall normalization of the CR and gamma-ray
distribution scales with the maximum acceleration efficiency at
structure formation shock waves. Following recent observations at
supernova remnants \cite{2009Sci...325..719H} as well as theoretical
studies \cite{2005ApJ...620...44K}, we adopt a realistic value of this
parameter and assume that 50\% of the dissipated energy at strong
shocks is injected into CRs while this efficiency rapidly decreases
for weaker shocks \cite{2007A&A...473...41E}.

@@

\subsection{Radiative processes}
In this section we outline the basics of inverse Compton (IC)
emission. As target radiation fields we consider both cosmic microwave
background (CMB) photons, and the light from stars and dust (SD). We
derive a semi-analytic model from which we can estimate the SD
spectral and spatial distributions as a simple function of cluster
mass.

>>>>

%NEW
\newcommand{\rmn}{\mathrm}
\newcommand{\bx}{{\bf x}}
\newcommand{\fg}{{F_\gamma}}
\newcommand{\sg}{{S_\gamma}}
\newcommand{\psf}{\theta_\rmn{res}}
\newcommand{\ph}{\rmn{ph}}
\newcommand{\eph}{E_\ph}
\newcommand{\nph}{N_\ph}
\newcommand{\vir}{\rmn{vir}}
\newcommand{\gal}{\rmn{gal}}
\newcommand{\sd}{\rmn{SD}}
\newcommand{\clu}{\rmn{cl}}
\newcommand{\sfe}{\rmn{sfe}}
\newcommand{\sub}{\rmn{sub}}
\newcommand{\msun}{M_\odot}
\newcommand{\ee}{E_\rmn{e}}
\newcommand{\stars}{\rmn{stars}}
\newcommand{\dust}{\rmn{dust}}
\newcommand{\lx}{L_\rmn{X}}
\newcommand{\s}{\rmn{s}}
\newcommand{\Kp}{\rmn{K}`}
\newcommand{\Ip}{\rmn{I}`}
\newcommand{\Js}{\rmn{J}^*}
\newcommand{\Jp}{\rmn{J}`}
\newcommand{\B}{\rmn{B}}
\newcommand{\bsub}{\B_\rmn{sub}}
\newcommand{\qCR}{q_{\rmn{CR}-\ensuremath{\pi^0}}}

%OLD
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\gev}{\rmn{GeV}}
\newcommand{\tev}{\rmn{TeV}}
\newcommand{\ev}{\rmn{eV}}
\newcommand{\dd}{\rmn{d}}
\newcommand{\mx}{\ensuremath{m_{\chi}}}
\newcommand{\ga}{\gamma}
\newcommand{\ngamma}{\ensuremath{N_{\gamma}}}
\newcommand{\ngammaj}{\ensuremath{N_{\gamma,j}}}
\newcommand{\sigmaannv}{\ensuremath{\langle\sigma v\rangle}}
\newcommand{\sigv}{\ensuremath{\sigma_v}}
\newcommand{\ic}{\rmn{IC}}
\newcommand{\egamma}{\ensuremath{E_{\gamma}}}
\newcommand{\dic}{\rmn{dIC}}
\newcommand{\fsr}{\rmn{FSR}}
\newcommand{\CR}{\rmn{CR}}
\newcommand{\rhos}{\ensuremath{\rho_s}}
\newcommand{\rs}{\ensuremath{r_s}}
\newcommand{\rvir}{r_{200}}
\newcommand{\mvir}{M_{200}}
\newcommand{\rhoc}{\ensuremath{\rho_c}}
\newcommand{\e}{\rmn{e}}
\newcommand{\eg}{E_\gamma}
\newcommand{\ir}{\rmn{ir}}

@@

Among the astrophysical systems which will be very interesting to
detect, and study, with coming gamma-ray detectors (Fermi, HESS,
MAGIC, VERITAS, and eventually large detectors like CTA) belong galaxy
clusters. The most promising directions in which to search for a
gamma-ray annihilation signal (from the annihilation process itself,
and also the accompanying bremsstrahlung and inverse Compton
components coming from charged particles produced in the annihilations)
are basically three:

@@

3. Galaxy clusters. This possibility has been less studied, however we
noted in a previous Letter \cite{2009PhRvL.103r1302P} that the are
certain advantages that work in favour of this possible target for
gamma-ray detection of dark matter annihilation.  Galaxy clusters
constitute the most massive objects in our Universe that are forming
today. This causes their DM subhalo mass function to be less affected
by tidal stripping compared to galaxy sized halos that formed long
ago.  The annihilation luminosities of the DM halo component for
e.g. the Virgo cluster and the Draco dwarf scales in a way (see
\cite{2009PhRvL.103r1302P}) that the ratio of gamma-ray luminosities
from the smooth components is around 4, in favour of Virgo. In
addition, there may be a further enhancement due to substructure,
which to a large extent should be unaffected by tidal disruption, at
least in the outer regions.  According to a recent estimate
\cite{2011MNRAS.410.2309G}, more massive haloes tend to have a larger
mass fraction in subhalos.  For example, cluster size haloes typically
have 7.5 per cent of the mass within $r_{200}$ in substructures of
fractional mass larger than $10^{-5}$, which is 25 per cent higher
than galactic haloes \footnote{All halo masses and length scales in
  this paper are scaled to the currently favored value of Hubble's
  constant, $H_0 = 70\, \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define the
  virial mass $\mvir$ and virial radius $\rvir$ as the mass and radius
  of a sphere enclosing a mean density that is 200 times the critical
  density of the Universe $\rho_{\rmn{cr}}$.}.

@@

\del{In this work we define the virial radius $\rvir$ of a halo to be
  the radius at which the mean density within is a factor $\Delta=200$
  times the critical density $\rhoc$ of the universe.}

@@

that the radius of the source $r=\sqrt{l^2+D^2-2 D l\cos\Psi}$, where

@@

We denote the total differential source function by $q_{\rm
  sum}(\eg,r)$, where we include contributions from five main
processes; leptophilic DM annihilating either indirectly through
$\mu^+/\mu^-$ to $\rmn{e}^+/\rmn{e}^-$ or directly to
$\rmn{e}^+/\rmn{e}^-$ pairs that inverse Compton upscatter background
photons (LP-IC), leptophilic DM emitting final state radiation
(LP-FSR), supersymmetric DM benchmark models where annihilating
neutralinos generate $\rmn{e}^+/\rmn{e}^-$ pairs that upscatter
background photons (BM-IC) and emit a continuum as well as final state
radiation (BM-Cont), and CR proton induced $\pi^0$:s decaying into
gamma-rays (CR-$\pi^0$). The source function is given by

@@

q_\rmn{sum} (\eg,r) = \qCR(\eg,r)+

@@

the differential CR to gamma-ray source function is denoted by
$\qCR(\eg,r)$ (see Sect.~\ref{sect:CRs} and \cite{2010MNRAS.409..449P}
for further details). The subscript $i$ runs over the gamma-ray
producing DM channels and the total differential boost factor for DM
is given by $\B_{\rmn{tot},i}(r,\sigv) =
\B_\sfe(\sigv)\,\B_{\sub,i}(r)$. It is the product of enhancement factors
from SFE $\B_\sfe(\sigv)$ (see Sect.~\ref{sect:LP}) and from
substructure enhancement over the smooth halo contribution
$\B_{\sub,i}(r) = 1+q_\sub(r)/q_{\rmn{sm},i}(r)$ (see
Sect.~\ref{sect:subst}). The DM source function from the smooth halo
for each process is written on the form:

@@

\frac{\dd \ngammaj}{\dd E_\gamma} \Gamma_j(r)\,,

@@

where the annihilation rate density is given by 

@@

\Gamma_j(r) = \frac{1}{2} \left[\frac{\rho(r)}{\mx}\right]^2 
\, \sigmaannv_j\,.

@@

Here the subscript $j$ runs over all kinematically allowed gamma-ray

@@

clusters. It can be considered as a special case of the more general
5-parameter profile:

@@

and the characteristic overdensity by 

@@

\label{eq:dens_ein}

@@

Navarro-Frenk-White (NFW) profile, but provide a better fit to recent
simulated high resolution DM halos \cite{2010MNRAS.402...21N}. It
should also be noted that recent detailed galaxy cluster observations
find a core like density profile with an inner slope of $\beta=0.6$,
where $\beta\gtrsim 1$ can be ruled out to a $>95$\% confidence
\cite{2011ApJ...728L..39N}. In Eq.~(\ref{eq:dens_ein}) we denote the
density where the profile has a slope of $-2$ by $\rho_{-2}$, and the
radius by $r_{-2}$. Assuming that all the flux from an NFW profile
originate from within the scale radius $\rs=\rvir/c$, and all the flux
from an Einasto profile originate within $r_{-2}=\rvir/c$, it follows
that $\rho_{-2} = \rhos/4$.

%However, this finding is hard to reconcile with simulations that
%typically prefere a steeper central profile.

\subsubsection{Substructures}
\label{sect:subst}
High-resolution DM only simulations of Milky Way (MW) type halos find
substantial amount of substructures in the periphery of DM halos,
while the substructures in the center are erased due to dynamical
disruption. Since the rate of which DM is annihilating depends on the
density squared, the resulting flux from substructures is boosted
compared to the smooth density distribution. Recent high resolution
simulations suggest a flux enhancement from substructures of the order
of ten up to a few 100 \cite{2008MNRAS.391.1685S, 2008JPhCS.125a2008K}
for a MW size halo. We use a double power-law function to
fit \footnote{Our approach of fitting the scaling behavior of
  $\rmn{L_{sub}}(<r)$ directly from numerical simulations
  self-consistently accounts for the radial dependence of the
  substructure concentration \cite{2008MNRAS.391.1685S}.} the
luminosity from the smooth component of substructures
(i.e. substructures within substructures are not included) inside
radius $r$ which is determined for the Aquarius simulations
\cite{2008MNRAS.391.1685S,2008Natur.456...73S}. Our best fit is given
by
\begin{equation}
  {\rm L_{sub}}(<r)= 0.76\,C_1\,C_2(\mvir)\,{\rm L_{200sm}}\,x^{0.95
    \displaystyle{x}^{-0.27}}\,,
  \label{eq:LSub}
\end{equation}
where ${\rm L_{200sm}}$ is the luminosity from the smooth halo without
substructures within $\rvir$ and $x=
\left(\frac{r}{\rvir}\right)$. The first normalization constant is
derived from the simulations:
\begin{equation}
  C_1=\left(\frac{M_{\rm res,sim}}{M_{\rm lim}}\right)^{0.226}\,,
   \label{eq:C1}
\end{equation}
where $M_{\rm res,sim}=10^5\msun$ is the mass of the smallest resolved
subhalos in the MW simulation, and $M_{\rm lim}$ is the theoretical
smallest mass of subhalos determined from free streaming length of DM
(CHECK). Here we assume that the power-law scaling relation in
Eq.~(\ref{eq:C1}) is valid down to the free streaming mass of DM
halos, which is conventionally in the cold dark matter universe
\cite{2001PhRvD..64h3507H,2005JCAP...08..003G} taken to be
$10^{-6}\msun$. Note that potentially the power-law could flatten
towards smaller mass scales although current simulations show no hits
of such a behavior and in addition we are approaching the asymptotic
behavior in the power spectrum on these scales(CHECK). For DM halos
more (less) massive than the MW we expect a larger (smaller) boost
from substructures, simply because of the larger (smaller) mass range
down to the minimum mass $M_\rmn{lim}$. We capture this halo mass
dependence with the second normalization in Eq.~(\ref{eq:LSub}):
\begin{equation}
C_2(\mvir)=\left(\frac{\mvir}{M_{\rm 200sim}}\right)^{0.226}\,,
 \label{eq:C2}
\end{equation}
where $M_{\rm 200sim}=1.9\times10^{12}\msun$ is the mass of the MW
halo in the simulation \cite{2008Natur.456...73S}.

The different density profiles have a large impact on the luminosity
from annihilating DM, although the details of the density profile can
be neglected compared to the dominating boost from substructures. In
Fig.~\ref{fig:radial_lum} we compare the radial dependence of the
accumulative luminosity from different smooth cluster density profiles
to the substructure boosted luminosity at different mass scales. We
recalculate the density contrast for the core like NFW density profile
$\rhos$ and rescale the concentration parameter in Eq.~(\ref{eq:cfit})
with $300/160$ \cite{2011ApJ...728L..39N} to account for the more
centralized scale radius in cluster with a softer cusp. The emission
coming from the core like NFW density profile ($\beta=0.6$) compared
to a cuspy ($\beta=1.0$) is about 30\% larger within $\rvir$. This
difference is build up within 10\% of $\rvir$ (i.e. close to
$r_\s$). Hence the slope of the central part of a cluster is
unimportant for the DM luminosity inside $\rvir$, instead the
important factor is the scale radius. The emission from an Einasto
density profile is about 50\% larger than the cuspy NFW profile in the
periphery of the cluster, where the difference is mainly build up at a
few percent of $\rvir$. The substructure boosted luminosities is
neglible in the center of halos, but integrated out to $\rvir$ the
boost is about; 10 for dwarf galaxies, $10^2$ for galaxies, and $10^3$
for galaxy clusters.
\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/dens.prof.eps}
\caption{\it Radial dependence of various luminosity
  contributions. The solid lines show the accumulative smooth
  luminosity from a cluster with the mass $\mvir=10^{15}\,\msun$ for
  three different density profiles; an Einasto profile with
  $\alpha=0.17$ (light blue), a cuspy NFW profile with $\beta=1.0$
  (dark blue), and a core NFW profile with $\beta=0.6$ (purple). The
  dashed lines show the accumulative luminosity from substructures for
  three different mass scales; an $\mvir=10^{15}\,\msun$ galaxy
  cluster (red), an $\mvir=10^{12}\,\msun$ galaxy (orange), and an
  $\mvir=10^{8}\,\msun$ dwarf galaxy (green). All luminosities have
  been normalized with the luminosity within $\rvir$ from a cuspy NFW
  profile. We have assumed the standard value for the limiting
  substructure mass of $M_\rmn{lim}=10^{-6}\,\msun$. Note the large
  expected boost from substructures in clusters ($\sim1000$), and the
  relative small boost in dwarf galaxies ($\sim20$).}
 \label{fig:radial_lum}
\end{figure}

Even though the bulk of substructures have been erased in the central
regions of DM halos, a cluster in projection has a significant
enhancement due to substructures at a radius of just a few percent of
$\rvir$. In Fig.~\ref{fig:radial_emis} we show the luminosity
dominated radius for substructures and smooth density profiles, both
for three different mass scales. It clearly shows how that the main
contribution from the substructures is coming from outer parts of the
DM halos, while the main contribution from the smooth density profiles
peaks around the scale radius. SHOW WHERE THE FLUX PEEKS WITH EQUATION
SEC DERIV
\begin{figure}
  \includegraphics[width=0.99\columnwidth]{figures/emissiv.sub.eps}
  \caption{\it Luminosity dominated radius for different mass
    scales. The solid lines show the emissivity from a cuspy
    $\beta=1.0$ NFW density profile for three different mass scales;
    an $\mvir=10^{15}\,\msun$ galaxy cluster (red), an
    $\mvir=10^{12}\,\msun$ galaxy (orange), and an
    $\mvir=10^{8}\,\msun$ dwarf galaxy (green). The dashed lines show
    the emissivity from substructures for the same three mass
    scales. All emissivities have been normalized with the luminosity
    within $\rvir$ from a cuspy NFW profile. We have assumed the
    standard value for the limiting substructure mass of
    $M_\rmn{lim}=10^{-6}\,\msun$.}
  \label{fig:radial_emis}
\end{figure}

\subsubsection{Cosmic-ray induced gamma-ray emission}
\label{sect:CRs}
On small astrophysical scales, such as supernovae remnants and
galaxies, there are convincing evidences of non-thermal
populations. Especially, in the MW, the cosmic rays are observed
directly as well as indirectly through radio, X-ray, and gamma-ray
emission. On larger scales of the order of few 100~kpc up to Mpc,
there are currently a vast number of observations of radio emission
coming from radio mini halos in the center of cooling flow clusters,
radio relics in the periphery of clusters \cite{2004rcfg.proc..335K}
as well as smooth giant radio halos that have been observed in more
than 50 clusters \cite{2003ASPC..301..143F,2008SSRv..134...93F}. This
type of emission is expected in clusters since the formation process
of a galaxy cluster is a very energetic processes that induces both
turbulence as well as frequently occurring merging and accretion
shocks which both are thought to accelerate large quantities of
relativistic non-thermal protons and radio emitting electrons to high
energies. Thus most of the observed radio emission in clusters is
expected to emerge from cosmic-ray electron induced radio synchrotron
emission. The precise origin of these electrons, in especially relics
and giant radio halos, is still not settled. However, due to the
smoothness of the extended emission and the observed power-law slope
it seems quite natural that they are of hadronic origin. Especially
since the long cooling time of cosmic ray protons (CRps) allows for
both an extended and large population of CRps to build up over the
cluster history \cite{1997ApJ...487..529B}.

When the CRps interact with ambient gas protons, it results in the
production of both charged and neutral pions that decay into
electrons, neutrinos, and gamma-rays. The production of these
secondaries depend both on the gas and CRp densities in the cluster,
where the CRp density roughly trace the gas outside the core regime and
is slightly enhanced in the center. This density scaling implies that
clusters are great targets for Cherenkov telescopes with a high
sensitivity for the central parts of close by clusters. Detecting the
cluster gamma-ray emission is crucial in this respect as it
potentially provides the unique and unambiguous evidence of CRp
populations in clusters through observing the $\pi^0$ bump at about
100~MeV in the spectra.

We adopt the spectrally and spatially universal gamma-ray model
developed by Pinzke \& Pfrommer \cite{2010MNRAS.409..449P} to estimate
the emission from decaying $\pi^0$:s that dominates over the inverse
Compton emission from primary and secondary electrons above 100~MeV in
clusters. The gamma-ray formalism was derived from high resolution
simulations of clusters of galaxies that included radiative
hydrodynamics, star formation, supernova feedback, and followed CRp
physics using a novel formulation that trace the most important
injection and loss processes self-consistently while accounting for
the CRp pressure in the equation of motion
\cite{2008A&A...481...33J,2007A&A...473...41E,2006MNRAS.367..113P}.
We note that the overall normalization of the CRp and gamma-ray
distribution scales with the maximum acceleration efficiency at
structure formation shock waves. Following recent observations at
supernova remnants \cite{2009Sci...325..719H} as well as theoretical
studies \cite{2005ApJ...620...44K}, we adopt a realistic value of this
parameter and assume that 50\% of the dissipated energy at strong
shocks is injected into CRps while this efficiency rapidly decreases
for weaker shocks \cite{2007A&A...473...41E}.

@@

\subsection{Radiative processes}
In this section we outline the basics of inverse Compton (IC)
emission. As target radiation fields we consider both cosmic microwave
background (CMB) photons, and the light from stars and dust (SD). We
derive a semi-analytic model from which we can estimate the spectral
and spatial distributions of SD as a simple function of cluster mass.
